---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```


# `TcGSA`

[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/TcGSA)](https://cran.r-project.org/package=TcGSA)
[![R-CMD-check](https://github.com/borishejblum/TcGSA/workflows/R-CMD-check/badge.svg)](https://github.com/borishejblum/TcGSA/actions)
[![Downloads](https://cranlogs.r-pkg.org/badges/TcGSA?color=blue)](https://www.r-pkg.org/pkg/TcGSA)

## Overview

`TcGSA` is a package which performs *Time-course Gene Set Analysis* from **microarray data**, and provide nice representations of its results.

On top of the CRAN help pdf-file, the following article explains what TcGSA is about:

>Hejblum, BP, Skinner, J, & Thiébaut, R (2015). Time-Course Gene Set Analysis for Longitudinal Gene Expression Data. *PLOS Computational Biology*, **11**(6):e1004310. [<doi: 10.1371/journal.pcbi.1004310>](https://doi.org/10.1371/journal.pcbi.1004310)



## Installation


TcGSA imports the `multtest` package which is not available on [CRAN](https://cran.r-project.org/), but is available on the [Bioconductor](https://www.bioconductor.org/packages/release/bioc/html/multtest.html) repository.
Before installing TcGSA, be sure to have this `multtest` package installed. If not, you can do so by running the following:

```{r bioc-install, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}
BiocManager::install("multtest")
```

The easiest way to get `TcGSA` is to install it from [CRAN](https://cran.r-project.org/package=TcGSA):
```{r CRAN-install, eval=FALSE}
install.packages("TcGSA")
```

or to get the development version from [GitHub](https://github.com/sistm/TcGSA):
```{r GitHub-install, eval=FALSE}
#install.packages("devtools")
devtools::install_github("sistm/TcGSA")
```

## Microarrays vs RNA-seq

`TcGSA` relies on a Gaussian assumption for the expression data, which is suitable for 
normalized microarray data. Due to their count and heteroskedastic nature, RNA-seq
data need to be handled differently and ***TcGSA cannot deal with RNA-seq data***.
For RNA-seq data, please have a look at the [Bioconductor package `dearseq`](https://bioconductor.org/packages/dearseq/) 
which incorporates similar functionality for analyzing RNA-seq data.

 -- Boris Hejblum
